---
title: "Using Randomization & Stratification to Overcome A Common Cause Confounder”"
subtitle: "Using Randomization & Stratification to Overcome A Common Cause Confounder”"
format:
  html: default
bibliography: references.bib
execute:
  echo: true
  eval: true
---


## Observational Data: A Misleading Victory for Doc Dreamy
```{r}
#| label: load-data-r
#| echo: true

# Load the observational data
patients_df <- read.csv('patients_data.csv')

# Display first few rows to understand the structure
cat("Number of patients:", nrow(patients_df), "\n")
head(patients_df)
```


Let's use data to figure out which surgeon performs better. @fig-plot-outcomes shows the post-surgical symptom score for 100 patients. Doc Duck's average post-surgical symptom score is 3.38 while Doc Dreamy's average is 2.8. Doc Dreamy performs better since lower scores indicate better outcomes. One might ask if the difference in average post-surgical symptom score is statistically significant. A two-sample t-test reveals a statistically significant difference (t = -2.317, p = 0.023). 

### Post-Surgical Outcomes by Patient
```{r}
#| label: fig-plot-outcomes
#| fig-cap: "Post-Surgical Symptom Score (lower is better) - Observed Data"
#| echo: true

library(ggplot2)

# Calculate means for each doctor
mean_dreamy <- mean(patients_df$post_surgical_score[patients_df$doctor_name == "Doc Dreamy"])
mean_duck <- mean(patients_df$post_surgical_score[patients_df$doctor_name == "Doc Duck"])

# Create the plot
ggplot(patients_df, aes(x = patient, y = post_surgical_score, color = doctor_name, shape = doctor_name)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_hline(aes(yintercept = mean_dreamy, color = "Doc Dreamy"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  geom_hline(aes(yintercept = mean_duck, color = "Doc Duck"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  annotate("text", x = max(patients_df$patient) * 0.95, y = mean_dreamy, 
           label = paste("○", round(mean_dreamy, 2)), 
           color = "#4E79A7", size = 4, fontface = "bold", hjust = 1) +
  annotate("text", x = max(patients_df$patient) * 0.95, y = mean_duck, 
           label = paste("△", round(mean_duck, 2)), 
           color = "#E15759", size = 4, fontface = "bold", hjust = 1) +
  scale_color_manual(values = c("Doc Dreamy" = "#4E79A7", "Doc Duck" = "#E15759"),
                     name = "Doctor") +
  scale_shape_manual(values = c("Doc Dreamy" = 16, "Doc Duck" = 17),
                     name = "Doctor") +
  labs(
    title = "Post-Surgical Symptom Score by Patient and Doctor",
    x = "Patient Number",
    y = "Post-Surgical Symptom Score (Lower is Better)",
    color = "Doctor",
    shape = "Doctor"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3)
  )

```


```{r}
#| label: load-data-randomized-r
#| echo: true

# Load the randomized data
patients_randomized_df <- read.csv('patients_data_randomized.csv')

# Display first few rows
cat("Number of randomized patients:", nrow(patients_randomized_df), "\n")
head(patients_randomized_df)
```


@fig-plot-outcomes-randomized shows the post-surgical symptom score for 100 patients under randomized assignment. Doc Duck's average post-surgical symptom score is 2.71 while Doc Dreamy's average is 3.46. Doc Duck performs better since lower scores indicate better outcomes. With randomization breaking the confounding relationship, we can now properly assess the causal effect. A two-sample t-test reveals a statistically significant difference (t = 2.734, p = 0.007).  **Not only is Dr. Dreamy not the better surgeon, he is actually the worse surgeon!**

```{r}
#| label: fig-plot-outcomes-randomized
#| fig-cap: "Post-Surgical Symptom Score (lower is better) for randomized patients."
#| echo: true

# Calculate means for each doctor in randomized data
mean_dreamy_rand <- mean(patients_randomized_df$post_surgical_score[patients_randomized_df$doctor_name == "Doc Dreamy"])
mean_duck_rand <- mean(patients_randomized_df$post_surgical_score[patients_randomized_df$doctor_name == "Doc Duck"])

# Create the plot
ggplot(patients_randomized_df, aes(x = patient, y = post_surgical_score, color = doctor_name, shape = doctor_name)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_hline(aes(yintercept = mean_dreamy_rand, color = "Doc Dreamy"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  geom_hline(aes(yintercept = mean_duck_rand, color = "Doc Duck"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  annotate("text", x = max(patients_randomized_df$patient) * 0.95, y = mean_dreamy_rand, 
           label = paste("○", round(mean_dreamy_rand, 2)), 
           color = "#4E79A7", size = 4, fontface = "bold", hjust = 1) +
  annotate("text", x = max(patients_randomized_df$patient) * 0.95, y = mean_duck_rand, 
           label = paste("△", round(mean_duck_rand, 2)), 
           color = "#E15759", size = 4, fontface = "bold", hjust = 1) +
  scale_color_manual(values = c("Doc Dreamy" = "#4E79A7", "Doc Duck" = "#E15759"),
                     name = "Doctor") +
  scale_shape_manual(values = c("Doc Dreamy" = 16, "Doc Duck" = 17),
                     name = "Doctor") +
  labs(
    title = "Post-Surgical Symptom Score (Randomized Assignment)",
    x = "Patient Number",
    y = "Post-Surgical Symptom Score (Lower is Better)",
    color = "Doctor",
    shape = "Doctor"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3)
  )

```

The key to handling a common cause confounder is to stratify by the common cause. In general, this means we examine the relationship between treatment and outcome within groups that share the same value of the confounder. In our example, this means we look at patients of similar severity and compare the outcomes of the surgeons for patients of similar severity. There are mathematically sophisticated ways to do this, but here we demonstrate it visually. Since patient number lacks inherent meaning, the following figure shows patient severity on the x-axis and post-surgical symptom score on the y-axis.  


```{r}
#| label: fig-plot-outcomes-severity
#| fig-cap: "Post-Surgical Symptom Score (lower is better) by Patient and Severity"
#| echo: true

# Calculate means for each doctor (using observational data)
mean_dreamy_sev <- mean(patients_df$post_surgical_score[patients_df$doctor_name == "Doc Dreamy"])
mean_duck_sev <- mean(patients_df$post_surgical_score[patients_df$doctor_name == "Doc Duck"])

# Create the plot
ggplot(patients_df, aes(x = severity, y = post_surgical_score, color = doctor_name, shape = doctor_name)) +
  # Add shaded region from -1 to 1 on x-axis
  annotate("rect", xmin = -1, xmax = 1, ymin = -Inf, ymax = Inf, 
           alpha = 0.15, fill = "gray", color = NA) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_hline(aes(yintercept = mean_dreamy_sev, color = "Doc Dreamy"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  geom_hline(aes(yintercept = mean_duck_sev, color = "Doc Duck"), 
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  annotate("text", x = max(patients_df$severity) * 0.95, y = mean_dreamy_sev, 
           label = paste("○", round(mean_dreamy_sev, 2)), 
           color = "#4E79A7", size = 4, fontface = "bold", hjust = 1) +
  annotate("text", x = max(patients_df$severity) * 0.95, y = mean_duck_sev, 
           label = paste("△", round(mean_duck_sev, 2)), 
           color = "#E15759", size = 4, fontface = "bold", hjust = 1) +
  scale_color_manual(values = c("Doc Dreamy" = "#4E79A7", "Doc Duck" = "#E15759"),
                     name = "Doctor") +
  scale_shape_manual(values = c("Doc Dreamy" = 16, "Doc Duck" = 17),
                     name = "Doctor") +
  labs(
    title = "Post-Surgical Symptom Score by Patient Severity",
    x = "Initial Patient Severity",
    y = "Post-Surgical Symptom Score (Lower is Better)",
    color = "Doctor",
    shape = "Doctor"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3)
  )

```
Looking at @fig-plot-outcomes-severity, the value of stratifying by severity is visually most obvious with initial patient severity scores between -1 and 1 (highlighted in the overlay). We chose this region because it shows substantial overlap between the two surgeons—both surgeons have considerable data within this range, making it an ideal region for comparing their performance. Outside of this range, there is less data available for making direct comparisons: Doc Duck sees more of the higher-severity patients (severity > 1), while Doc Dreamy sees more of the lower-severity patients (severity < -1).

Within the highlighted region, Doc Dreamy's blue circles tend to be higher (worse) than Doc Duck's red triangles for these patients. It is also obvious that Doc Duck is seeing the more severe patients overall, as his red triangles are generally to the right of Doc Dreamy's blue circles on the x-axis. This is exactly what we would expect if patient severity were a common cause of surgeon choice and surgical outcomes. Despite the aggregate statistics pointing to Doc Dreamy having lower (better) post-surgical symptom scores on average, our visual analysis of this overlapping region leads us to conclude that Doc Duck is actually the better surgeon—a conclusion that would be obscured if we only looked at the means without stratifying by severity.


